On Optimality of Finite Element Integration

### 2016-02-15 TOMS_Paper/assembly.tex
Rewrote algo talking about cycles etc

\usepackage{lineno}
\usepackage{xfrac}

@@

\item Restructure the expression based on the structure of $G$ and a set of rewrite rules $<$\textit{precondition}: \textit{action}$>$, to be applied in the order given below.
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace t^1, ..., t^n \rbrace \subset T$ adjacent to $s \in S$, $n > 1$: take $s$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $s (...t^1 + ... + ...t^n)$}.
\item Only one terminal $t$ adjacent to $s$: take an arbitrary non-terminal symbol $s^{nt}$ adjacent to $s$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $s (...t + ...s^{nt})$}. 
\item $T = \emptyset$, $C = [s, ..., v, ..., s]$ is a cycle, $v$ has highest degree in $C$, $\lbrace s^1, ..., s^n \rbrace$ adjacent to $v$: take $v$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $v (...s^1 + ... + ...s^n)$}.
\end{enumerate}
Rule (C) may expose further hits for rules (A) and (B). Rules are therefore applied cyclically until no more hits are possible.\\
\textit{Note: the last two steps only impact bilinear forms, since otherwise $E = \emptyset$.}

@@

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)). We then have (step (3)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, although no rewrite rule is applicable since $S = T$

@@

After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, a set of rewrite rules are applied. We here discuss why applying, in rigorous order, rules (A), (B), and (C), leads to the optimum.

Rules (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid no matter how Strategy~\ref{strategy:ii} is scheduled. In particular:
\begin{description}
\item[Rule (A)] The terminals $\lbrace t^1, ..., t^n \rbrace$ can only be grouped through the factorization of $s$ (by definition of terminal), so this rule has no side effects.
\item[Rule (B)] Since $s^{nt}$ is not a terminal, there must exist at least one edge between $s^{nt}$ and a generic vertex $v$. Let us consider the alternative factorization $s^{nt}(...s + ... + ...v)$. If $v$ were a terminal, we observe that setting $v = t$ and $s = s^{nt}$ would simply expose another hit for rule (B). Otherwise ($v$ non terminal), we would have $s$ now appearing in two different sub-expressions ($(...s + ... + ...v)$ and $(...s\ t)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $s$. From this, we evince that applying (C) prior (B) may prune the optimum from the search space.
\end{description}
To reach the optimum, rule (C) discards all those potential factorizations that can be rather ``aggregated'' with \textit{essential} factorizations. A factorization is essential if a vertex is in a cycle. Being in a cycle is the fundamental condition to

Having higher degree simply means grouping more symbols (i.e., avoiding more products);

%Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules. We here demonstrate why this is a step forwards towards the optimum. Note that, at this point, non flop-decreasing factorizations have already been pruned (step (5)). 
%
%Instances of 
%\begin{description}
%\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
%\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ were a terminal, we see that setting $V = T$ and $S = S^{nt}$ would simply expose another hit for rule (B). Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ... + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$. From this, we evince that applying rules (C) and (D) prior to rule (B) may prune the optimum from the search space.
%\end{description}
%
%Rule (C) is the critical one. Let $\mathbb{A}$ be an acyclic sub-graph rooted in $S$, and let $W$ be the neighbor of $S$ in $\mathbb{A}$. Since $\mathbb{T} = \emptyset$, $S$ is also the root a second sub-graph $\mathbb{C}$. The non-trivial case in when $\mathbb{C}$ is cyclic. If any vertex in $\mathbb{C}$ were factorized prior to $S$, the product $S\ W$ could eventually appear in the transformed expression. Conversely, is Strategy~\ref{strategy:ii} were first applied to $S$, $W$ and one or more symbols in $\mathbb{C}$ would be grouped in the same sub-expression and hoisted. 
% 
%Finally, note that the algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
\end{enumerate}
\end{proof}

@@

The rest of the section is organized as follows: we first provide insights into the overall outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we explain, in detail and individually for each form, the performance results obtained.

\paragraph{High level view}
The main observation is that our transformation strategy does not always guarantee minimum execution time. In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4, nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, memory constraints are critical, particularly the transformation strategy adopted when exceeding $T_H$. We will later thoroughly elaborate on all these aspects.

\paragraph{Autovectorization}
The discretizations employed result in inner loops and basis function tables of size multiple of the machine vector length. This, combined with the chosen compilation flags, promotes autovectorization in the majority of code variants. An exception is \texttt{quad} due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; manual inspection of the compiled code suggests, however, that block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have similar structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could evince from compiler reports and manual inspection of assembly code.

@@

\section{Conclusions}
\label{sec:conclusions}
We have developed transformation algorithms and a theory of optimality for finite element integration. The implementation was carried out in COFFEE, a compiler fully integrated with the Firedrake framework, and therefore made available to an increasingly significant user base. Performance results show the effectiveness of our methodology. There are still several open problems. The main one is an elaborate classification of forms for which optimality can be reached. Secondly, we would like to extend our methodology to classes of loops arising in spectral methods, in which interaction with low level optimization becomes even more critical due to the larger working sets.

>>>>

\usepackage{lineno}
\usepackage{xfrac}
\usepackage{amssymb}

@@

\item Use rewrite rules to restructure the expression. Let $V = \lbrace v^1, ..., v^n \rbrace$ be the vertices adjacent to $s \in S$; the preconditions for the three rewrite rules (A), (B), and (C) are:
\begin{enumerate}[label=(\Alph*)]
\item $A \subsetneq V$ is a set of vertices not standing on cyclic paths, $A \neq \emptyset$
\item $s$ is in a cycle $C$ and has highest degree among all vertices in $C$
\item $\lbrace v^1, ..., v^{n-1} \rbrace$ are leaves of a tree, $p = v^n$ is the parent of $s$ (unless $s$ is the root)
\end{enumerate}
If a precondition is met, we take $s$ and apply Strategy~\ref{strategy:ii} (i.e., we factorize $s$ and perform code motion), and we remove all edges incident to $s$. The preconditions are checked cyclically (first all (A)s, then all (B)s, with (B) that may expose further hits for (A), and finally all (C)s) until no more hits are possible (i.e., until $S = T$).  \\
\textit{Note: steps (4) and (5) only impact bilinear forms, since otherwise $E = \emptyset$.}

@@

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)). We then have (step (2)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, although no rewrite rule (step (5)) is applicable since $S = T$

@@

After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, a set of rewrite rules are applied. We here discuss why applying in order rules (A), (B) and (C) leads to the optimum.

Rule (A) is the critical one because it coordinates factorization at the junction between a cyclic and an acyclic components. It exploits the observation that a vertex outside of cyclic paths may be ``aggregated'' within one or more ``potentially unavoidable'' factorizations, thus saving on the final number of multiplications. This is quite evident for terminals (for which there is at most one way of being factorized), although not so obvious for vertices of arbitrary degree. Let $a$ be a vertex outside of cyclic paths, and let $V = \lbrace v^1, ..., v^n \rbrace$ be the set of vertices adjacent to $a$. Consider the non-trivial case in which at least one vertex $s \in V$ is within a cycle. If $a$ and $s$ were factorized separately, we could end up with the two sub-expressions $a(...v^1 + ... s + ...v^n)$ and $s(...)$. The spatial sharing induced by $s$ would make the synthesis sub-optimal. Conversely, if $s$ is factorized such that a single sub-expression $s(...a + ...)$ will be produced, the operation count is at worst unchanged.

Rule (B) acts within cycles by iteratively picking the vertex with highest degree. Since the assumption is that there are no more hits for rule (A), the problem here is as simple as selecting the minimum number of vertices whose incident edges would cover the whole cycle. With a very similar concept, rule (C) schedules the remaining (disconnected) sub-trees.
\end{enumerate}
\end{proof}

@@

The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we motivate, for each form, the performance results obtained.

\paragraph{High level view}
Our transformation strategy does not always guarantee minimum execution time. In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4, nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on the actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.

\paragraph{Autovectorization}
We discretized our problems such that the inner loop sizes would be a multiple of the machine vector length. This ensured autovectorization in the majority of code variants\footnote{The compilation flag \texttt{xHost} helped the Intel compiler generating efficient vector code. We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}. The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.

@@

\section{Conclusions}
\label{sec:conclusions}
We have developed transformation algorithms and a theory of optimality for finite element integration. The implementation was carried out in COFFEE, a compiler fully integrated with the Firedrake framework, and therefore made publicly available. Performance results suggest the effectiveness of our methodology. There are still several open problems. Among them, a classification of forms for which optimality can be reached and a generalization of the methodology to other loop nests, for instance those arising in spectral methods.
